---
title: "Exploring Diabetes Risk Through a Cell-Based Simulation Model"
author: "Joshua Edefo"
email: "edefojoshu2000@yahoo.com"
date: "2025-02-15"
output: github_document
---
The cell-based model simulates diabetes progression in a synthetic population of 1000 individuals over 10 years, factoring in age, gender, diet, and exercise. Each person is assigned a risk level (Low, Moderate, or High) based on these factors. Transition probabilities vary by risk level, affecting the likelihood of staying healthy or progressing to Pre- diabetes, Diabetes, or complications. After 10 years, most individuals (894) remain healthy, while fewer progress to Pre-diabetes (78) or Diabetes (27), and only 1 person experiences complications, reflecting the long-term impact of lifestyle and demographics on diabetes progression.

Libraries
```{r a, message=FALSE}
library(usethis)
library(dplyr)
```

Cell Based Model

```{r b}
## Set directory
setwd("C:\\Users\\joe62\\OneDrive - Aberystwyth University\\Apps\\Desktop\\Destop Folder")

## Set parameters
set.seed(123)  # For reproducibility
years <- 10    # Number of years to simulate
population_size <- 1000

## Generate a synthetic population
population <- data.frame(
  id = 1:population_size,
  age = sample(30:80, population_size, replace = TRUE),
  gender = sample(c("Male", "Female"), population_size, replace = TRUE),
  diet = sample(c("Healthy", "Moderate", "Unhealthy"), population_size, replace = TRUE, prob = c(0.3, 0.4, 0.3)),
  exercise = sample(c("High", "Moderate", "Low"), population_size, replace = TRUE, prob = c(0.3, 0.4, 0.3)),
  state = "Healthy"  # Initial state
)

## Function to assign risk category based on demographics, diet, and exercise
assign_risk <- function(age, gender, diet, exercise) {
  risk <- 0
  if (diet == "Unhealthy") risk <- risk + 2
  if (exercise == "Low") risk <- risk + 2
  if (age > 50) risk <- risk + 1
  if (gender == "Male") risk <- risk + 1
  
  if (risk <= 1) return("Low")
  if (risk == 2 | risk == 3) return("Moderate")
  return("High")
}

## Assign risk levels using vectorized approach
population$risk <- mapply(assign_risk, population$age, population$gender, population$diet, population$exercise)

## Transition probabilities based on risk category
transition_probs <- list(
  Low = c(Healthy = 0.98, PreDiabetes = 0.02, Diabetes = 0, Complications = 0),
  Moderate = c(Healthy = 0.90, PreDiabetes = 0.08, Diabetes = 0.02, Complications = 0),
  High = c(Healthy = 0.75, PreDiabetes = 0.15, Diabetes = 0.08, Complications = 0.02)
)

## Function to simulate state transition based on risk
simulate_transition <- function(risk) {
  probs <- transition_probs[[risk]]
  states <- c("Healthy", "PreDiabetes", "Diabetes", "Complications")
  new_state <- sample(states, 1, prob = c(probs["Healthy"], probs["PreDiabetes"], probs["Diabetes"], probs["Complications"]))
  return(new_state)
}

## Simulate over time
for (year in 1:years) {
  population$state <- sapply(population$risk, simulate_transition)
}

## Summary of final states
state_summary <- table(population$state)
print(state_summary)

```

summary of the model output
After 10 years, the majority of individuals (894) remained healthy. 
A smaller number developed PreDiabetes (78) or Diabetes (27). 
Only one individual progressed to complications,


Session information
```{r cE}
sessionInfo()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
